---
title: "Numbers in Colonial Maya Texts"
author: "Alex LaPrevotte"
date: "30 April 2026"
format: gfm
---

```{r setup, include=FALSE}
library(tidyverse)
library(pdftools)
```

**I have opted to add my new code, as I go, to my existing quarto document.**

My goal for processing these data is to get the text into a format I can work with. The data, available as part of "Los títulos de ebtún : transcripción, traducción y análisis histórico," the PhD dissertation of Dr. Julien Machault, which can be found [here](https://tesiunamdocumentos.dgb.unam.mx/ptd2025/abr_jun/0869324/Index.html), are currently in a PDF, broken into chunks of five lines:

* a transcription of the original colonial Maya text (in original orthography)
* a transcription of the text with modern orthography
* a morphological breakdown
* a morphological gloss
* a Spanish translation.
  
I am working with a 200-page excerpt from this PDF (pages 301-500) and I am attempting, as much as possible, to account for situations that do not arise in the current data within the code, so it could be applied to a larger data set. I anticipate my working data, pre-analysis, will take the form of a data frame with two columns (morphological breakdown, morphological gloss) and many rows, one per word. I plan to eliminate the original orthography, modern orthography, and Spanish translation because they contain special characters, do not correspond well with the other lines in number of words, and the only way to correlate the Spanish translations with the Maya would be manually, due to differences in word order and agglutination. 

## Reading in the data

```{r data-read-in}
# read in the text with each page as a row
ebtun_text <-pdf_text("data/ebtun_sample.pdf")

# lines to rows
ebtun_text <- strsplit(ebtun_text, "\n")

# combine all pages into one vector
ebtun_text <- unlist(ebtun_text)

# currently 1 column, 7695 rows
```

## Combining morphological breakdowns into single units

```{r fixing-split-words}
# eliminating multiple spaces and spaces before and after hyphens
# this should keep morphological breakdowns and glosses together
ebtun_text <- ebtun_text |>
  str_replace_all("[ \\t]+-", "-") |> 
  str_replace_all("-[ \\t]+", "-") |> 
  str_squish()                        

# eliminating null characters
ebtun_text <- str_replace_all(ebtun_text, "\\bø\\b", "")

# convert to dataframe
ebtun_data <- data.frame(line = ebtun_text)

# still 1 column, 7695 rows
```

## Cleaning up the rows

```{r row-cleaning}
# convert everything to lower case
ebtun_data <- ebtun_data %>% mutate(across(where(is.character), tolower))

# eliminating incomplete chunks, parts entirely in Spanish, document titles, and footers
# eliminating chunks where one or more lines goes onto a second row, disrupting the 5-line chunk format
ebtun_data <- ebtun_data[-c(1, 252, 598:618, 671:676, 679:684, 885:890, 1080:1085, 1117:1122, 1133:1136, 1141:1185, 1307:1312, 1401:2043, 2079, 2558:2591, 2616, 2882:2891, 2919, 3226, 3587:3592, 3760, 3906:3981, 4748, 4826, 5384:5419, 5440, 6050, 6454:6493, 6891, 7261, 7274:7282, 7295:7315, 7401, 7689:7692), ]

# again, we make it a data frame
ebtun_data <- data.frame(line = ebtun_data)

# eliminate rows that only include numbers, and spaces as well as rows that start with "doc.", "verso", or "sigue en"
ebtun_data <- ebtun_data %>%
  filter(
    !grepl("^\\s*$", .data$line) &
    !grepl("^\\s*\\d{,3}\\s*$", .data$line) &
    !grepl("^\\s*(verso|doc\\.|sigue en)", .data$line)
  )

# make it a dataframe yet again
ebtun_data <- data.frame(line = ebtun_data, stringsAsFactors = FALSE)

# currently 1 column, 3965 rows
```

## Breaking the data into chunks

```{r data-to-chunks}
# my goal here is to break the data back down into the 5-line chunks that were the basis of the PDF formatting

# every 5 lines to a group
ebtun_grouped <- ebtun_data |>
  mutate(group = ceiling(row_number() / 5))

# each group to a dataframe
ebtun_chunks <- split(ebtun_grouped, ebtun_grouped$group)

# currently 793 data frames, each two columns (line and group number) and five rows
```

## Removing rows not used in analysis

```{r whittling-rows}
# eliminating rows 1 (original text), 2 (modern orthography), and 5 (Spanish translation) from each chunk
ebtun_chunks <- lapply(ebtun_chunks, function(df) {
  df[3:4, , drop = FALSE]
})
```

## Split chunks by word

```{r split-by-word}
# function: split chunk into words
  # split each line by spaces
  # find max words in the chunk and add NAs to standardize columns
  # combine into data frame
split_chunk <- function(df_chunk) {
  split_words <- str_split(df_chunk$line, "\\s+", simplify = FALSE)
  max_words <- max(lengths(split_words))
  split_words_padded <- lapply(split_words, function(x) {
    length(x) <- max_words
    x
  })
  df_words <- as.data.frame(do.call(rbind, split_words_padded), stringsAsFactors = FALSE)
  return(df_words)
}

# use function on chunks
ebtun_chunks <- lapply(ebtun_chunks, split_chunk)

#currently 2 rows, 4491 columns (793 chunks)
```

## NA wrangling

```{r NA-wrangling}
# which chunks have NAs? (visual check)
na_counts <- sapply(ebtun_chunks, function(df) sum(is.na(df)))
#na_counts

# eliminate chunks with NAs
ebtun_chunks_nona <- ebtun_chunks[!sapply(ebtun_chunks, function(df) any(is.na(df)))]

# check to make sure number of chunks has decreased
length(ebtun_chunks_nona)

# down to 770 chunks
```

## Chunk concatenation

```{r chunk-concatenation, message=FALSE}
# number of rows
num_rows <- nrow(ebtun_chunks_nona[[1]])

# concatenate rows by chunk (this takes a while)
ebtun_combined <- map_dfr(1:num_rows, function(i) {
  rows <- lapply(ebtun_chunks_nona, function(df) df[i, , drop = FALSE])
  suppressMessages(
    bind_cols(rows, .name_repair = "minimal")
  )
})

#drop the column names
colnames(ebtun_combined) <- NULL
```

## Data transposition

```{r transpose-data}
# turn those rows into columns
ebtun_long <- t(ebtun_combined)

# convert back to data frame
ebtun <- as.data.frame(ebtun_long)

# currently 2 columns (morphological breakdown and morphological gloss), 4311 rows
```

## Final data cleaning

```{r final-data-cleaning}
# removing leading and trailing spaces
ebtun$V1 <- str_trim(ebtun$V1)
ebtun$V2 <- str_trim(ebtun$V2)

# remove rows containing cells that are just ellipses (either the ellipses character or three consecutive periods)
ebtun <- ebtun |>
  filter(if_all(everything(), ~ !str_trim(.) %in% c("...", "…")))

#naming the columns because I'm about to add more and it seems responsible
ebtun <- ebtun |>
  rename(
    breakdown = V1,
    gloss = V2
  )

# final form of "found" data
str(ebtun)

# currently 2 columns (morphological breakdown and morphological gloss), 3988 rows
```

## Analysis

```{r analysis, message=FALSE}
# reading in lists of Spanish and Maya numbers (textual)
esp_numbers <- read_lines("data/esp_numbers.txt")
maya_numbers <- read_lines("data/maya_numbers.txt")

# fix encoding for accents
esp_numbers <- iconv(esp_numbers, from = "latin1", to = "UTF-8")
maya_numbers <- iconv(maya_numbers, from = "latin1", to = "UTF-8")

# define numbers for gloss column (numerals or spanish numbers as morph segments)
number_pattern <- paste0(
  "(^|-)",
  "(",
  "\\d+",
  "|",
  paste(esp_numbers, collapse = "|"),
  ")",
  "($|-)"
)

# find solo numbers in gloss column
ebtun <- ebtun |>
  mutate(
    gloss_number = str_detect(gloss, number_pattern)
  )

# find numerals in breakdown column
ebtun <- ebtun |>
  mutate(
    numeral = str_detect(str_trim(breakdown), "(^|-)[0-9]+($|-)"))

# find Maya and Spanish numbers in breakdown column
ebtun <- ebtun |>
  mutate(
    maya_number =
      str_detect(breakdown, paste(maya_numbers, collapse = "|")) &
      gloss_number,
    spanish_number =
      str_detect(breakdown, paste(esp_numbers, collapse = "|")) &
      gloss_number
  )

str(ebtun)
#currently 6 columns, 3988 rows

# counting how many numerals, Maya numbers, and Spanish numbers are in the data
number_counts <- data.frame(
  type = c("numeral", "maya_number", "spanish_number"),
  count = c(
    sum(ebtun$numeral, na.rm = TRUE),
    sum(ebtun$maya_number, na.rm = TRUE),
    sum(ebtun$spanish_number, na.rm = TRUE)
  )
)

number_counts

view(ebtun)
```



```{r session-info}
sessionInfo()
```